\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-08-01 mers-structure.tex
Add scripts, some data, last hi-res figures, some text changes.

@@

We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{mers_epi}).
For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters (Figure \ref{mers_epi}).
Under unbiased sequencing only 0.6\% of simulations fit our phylogenetic observations, while 2.7\% and 2.7\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
Correspondingly, we estimate 10\% support for a model with bias level 1.0, 45\% support for a model with bias level 2.0, and 45\% support for a model with bias level 3.0.
Model averaging would suggest plausible $R_0$ values between \tbc{XXX} and \tbc{XXX}.

Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{mers_epi}).
The median number of cross-species introductions observed in simulations matching empirical data without bias are 339 (95\% percentiles 255--431).
These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
Model averaging would suggest plausible numbers of introductions between \tbc{XXX} and \tbc{XXX}.
Our results suggest a large number of unobserved MERS primary cases.
Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
In these analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).

@@

% \begin{figure}[h]
% \centering
%
% 	\caption{\textbf{Pairwise co-occurence matrix for human clusters.}
% Heatmap shows the posterior probability that a pair of tips from trees of genomic fragments fall within the same clade - tips from fragment 1 are on the x axis, tips from fragment 2 are on the y axis.
% Tips are ordered by their appearance in tree of genome fragment 2 (positions from nucleotide 21000 onwards) reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
% Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
% 	}
% 	\label{coocurrence}
% \end{figure}
\begin{figure}%
    \centering
    \includegraphics[width=0.95\textwidth]{figures/mers_recombinant_features.png}
    \caption{\textbf{Recombinant features of MERS-CoV phylogenies.}
    A) Heatmap showing the posterior probability that viruses from trees of different genomic fragments belong to the same introduction event.
    Genomic fragment 1 represents nucleotide positions 1 to 21000 and genomic fragment 2 represents nucleotide positions 21001 to \tbc{XXX}.
    The same set of viruses are arrayed on the $x$-axis and on the $y$-axis; the $x$-axis shows identity of these viruses along genomic fragment 1 and the $y$-axis shows identity of these viruses along genomic fragment 2.
    Viruses are ordered by their appearance in tree of genomic fragment 2 reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
    Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
    B) Phylogenies derived from genomic fragment 1 (left) and genomic fragment 2 (right), reduced to just the human tips.
  	Identical taxa are connected via coloured lines with colour indicating vertical position in the tree on the right.
  	Branches are coloured by inferred ancestral host state (human in blue, camel in orange).
  	Human clusters in blue change phylogenetic positions between the trees together, with little incongruence within clusters.
  	This is evidence for recombinant viruses generated in the camel reservoir entering human populations.
    }
    \label{recombination}
\end{figure}

@@

Our attempt at controlling the confounding effects of different sampling regimes between humans and camels revealed a troubling pattern, however.
We find that this camel-like dataset results in poor MCMC performance, where independent chains reach stationary distributions which are not necessarily the same.
Out of five runs, two chains converged to one stationary distribution, two to another and the last one to a third.
Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.
\tbc{This entire paragraph should be moved to Methods. Too much detail for Results. Below just say that ``Multiple MCMC chains showed similar patternso of $N_e \tau$ through time.''}

@@

\subsection*{MERS-CoV epidemiology}
%\lmc{Detecting?}
Understanding population structure yields important insight into any biological system.
It also often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
\tbc{This seems way overly broad to me. I much preferred the Introductions take on genomic epi revealing important population structure. Just revise to keep this focused on pathogen transmission. Identification of reservoirs, source-sink, all very important.}
Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula by using sequence data collected from both hosts (174 from humans and 100 from camels).

We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV genomic data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (Figures \ref{mcc}, \ref{exploded}).
This inference succeeds under different choices of priors for unknown demographic parameters (Figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (Figure \ref{mers_epi}).
From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (Figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (Figure \ref{mers_epi}).
We also looked at potential seasonality in MERS-CoV spillover into humans.
Our analyses indicated a period of three months where the odds of a sequenced spillover event are increased, with timing consistent with an enzootic amongst camel calves (Figure \ref{seasonality}).
As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (Figure \ref{mers_epi}).

\subsection*{Transmissibility differences between zoonoses and pandemics}
Severe acute respiratory syndrome (SARS) coronavirus, a Betacoronavirus like MERS-CoV, caused a serious epidemic in humans in 2003, with over 8000 cases and nearly 800 deaths.
Since MERS-CoV was also able to cause significant pathogenicity in the human host it was inevitable that parallels would be drawn between MERS-CoV and SARS-CoV at the time of MERS discovery in 2012.
Although we describe the epidemiology of MERS-CoV from sequence data, indications that MERS-CoV has poor capacity to spread human-to-human existed prior to any sequence data.
SARS-CoV swept through the world in a short period of time, but MERS cases trickled slowly and were restricted to the Arabian Peninsula or resulted in self-limiting outbreaks outside of the region, a pattern strongly indicative of repeat zoonotic spillover.
Infectious disease surveillance and control measures remain limited, so much like the SARS epidemic in 2003 or the H1N1 pandemic in 2009, zoonotic pathogens with $R_{0}>1.0$ are probably going to be discovered after spreading beyond the original location of spillover.
\tbc{I like the basic approach, this last segue could be cleaner however.}

@@

\section*{Methods}
\subsection*{Sequence data}
All MERS-CoV sequences were downloaded from GenBank and accession numbers are given in Table \ref{sequences}.
\tbc{Mention what's going on with accession-less sequences.}
Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
\tbc{This should be obvious in Table \ref{sequences}. Some strains should be multiple accessions. However, I don't see any.}
\tbc{What we the criteria for inclusion? How many bases were necessary to be considered a `genome'?}
Sequences were annotated with available collection dates and hosts, designated as camel or human.
Protein coding sequences were extracted and concatenated.
\tbc{Leaving how many bases?}
\tbc{How were they aligned?}
The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections (Table \ref{sequences}).

@@

MultiTypeTree module \citep{vaughan_efficient_2014} was used in BEAUti v2.4.3 \citep{bouckaert_beast_2014} to specify a structured coalescent model with two demes -- humans and camels.
This model estimates 4 parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
All analyses were run across ten independent Markov chains (MCMC runs).
\tbc{What was burn-in? How many steps in the MCMC? How was convergence assessed? How were chains combined?}
To confirm that demographic inference was derived from the data rather than the prior we set up an additional ten independent analyses where the mean of the exponential distribution prior for migration rate was set to 10.0.

As a secondary analysis, we also considered a scenario where alignments split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
All analyses were set to run for 200 million states, subsampling every 20 000 states.
Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in for single fragment analyses and 20\% for two-fragment analyses.
Chains not converging after 200 million states were discarded.
For single fragment analysis with migration rate prior set to exponential with mean 1.0 there were three chains out of ten that did not converge and two for the analysis where the migration rate prior was an exponential with mean 10.0.
For the two fragment structured coalescent analyses three chains out of ten failed to converge.

\subsection*{Demographic inference of MERS-CoV in the reservoir}
In order to infer the demographic history of MERS-CoV in camels we used the results of structured coalescent analyses to identify introductions of the virus into humans.
The oldest sequence from each cluster introduced into humans was used to represent a random draw from the diversity of MERS-CoV circulating in camels.
These sequences were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity (N$_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.%\lmc{already reported above. Why not remove?} \gdc{Never hurts to repeat.}

@@

  116 & Al-Hasa\_7\_2013 & KF600623 & human & 2013-05-01 \\
  117 & Al-Hasa\_8\_2013 & KF600618 & human & 2013-05-01 \\

@@

  119 & Al-Hasa\_11\_2013 & KF600629 & human & 2013-05-03 \\

@@

  121 & Al-Hasa\_14\_2013 & KF600615 & human & 2013-05-08 \\

@@

  126 & Al-Hasa\_23\_2013 & KJ156860 & human & 2013-05-13 \\

@@

  134 & Taif\_2\_2013 & KJ156896 & human & 2013-06-12 \\
  135 & Taif\_3\_2013 & KJ156938 & human & 2013-06-13 \\
  136 & Al-Hasa\_26\_2013 & KJ156882 & human & 2013-06-18 \\
  137 & Al-Hasa\_27\_2013 & KJ156943 & human & 2013-06-19 \\
  138 & Al-Hasa\_28\_2013 & KJ156887 & human & 2013-06-22 \\
  139 & Riyadh\_6\_2013 & KJ156879 & human & 2013-07-02 \\

@@

  141 & Riyadh\_7\_2013 & KJ156937 & human & 2013-07-15 \\
  142 & Riyadh\_8\_2013 & KJ156880 & human & 2013-07-17 \\

@@

  145 & Asir\_2\_2013 & KJ156863 & human & 2013-08-05 \\
  146 & Riyadh\_11\_2013 & KJ156946 & human & 2013-08-06 \\
  147 & Riyadh\_12\_2013 & KJ156926 & human & 2013-08-08 \\
  148 & Riyadh\_13\_2013 & KJ156888 & human & 2013-08-13 \\

@@

  150 & Hafr-Al-Batin\_4\_2013 & KJ156931 & human & 2013-08-25 \\
  151 & Hafr-Al-Batin\_5\_2013 & KJ156951 & human & 2013-08-25 \\
  152 & Riyadh\_17\_2013 & KJ156918 & human & 2013-08-26 \\

@@

  154 & Riyadh\_10\_2013 & KJ156891 & human & 2013-09-05 \\
  155 & Madinah\_3b\_2013 & KJ156916 & human & 2013-09-11 \\

@@

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_prior.png}
	\caption{\textbf{Result are robust to prior choice.}
Negligible flow of MERS-CoV lineages from humans into camels is recovered regardless of prior choice.
Plots show the 95\% highest posterior density for the estimated migration rate from the human deme into the camel deme (blue) and \textit{vice versa} (orange).
Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (bottom) or 10.0 (top).
%\tbc{I think this would be more clear if the left and right sides shared the same x-axis. As it stands it looks at first glance like there's different priors for camel-to-human compared to human-to-camel. It's okay if the blue human-to-camel rates gets smushed up against the left of the plot.}
	}
	\label{prior}
\end{figure}

>>>>

Although we find evidence of seasonality in zoonotic spill over timing, no such relationship exists for sizes of human sequence clusters (\textit{i.e. post-spillover clades})(?).
This is entirely expected, since little seasonality in human behaviour that could facilitate MERS-CoV transmission is expected following an introduction.
Similarly, we do not observe any trend in human sequence cluster sizes over time, suggesting that MERS-CoV outbreaks in humans are neither growing nor shrinking in size.
This is not surprising either, since MERS-CoV is a camel virus that experiences little to no selective pressure to improve transmissibility between humans.

@@

We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{mers_epi}).
For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters (Figure \ref{mers_epi}).
Under unbiased sequencing only 0.6\% of simulations fit our phylogenetic observations, while 2.7\% and 2.7\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
Correspondingly, we estimate 10\% support for a model with bias level 1.0, 45\% support for a model with bias level 2.0, and 45\% support for a model with bias level 3.0.
Model averaging would suggest plausible $R_0$ values between 0.57 and 0.91.

Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{mers_epi}).
The median number of cross-species introductions observed in simulations matching empirical data without bias are 339 (95\% percentiles 255--431).
These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
Model averaging would suggest plausible numbers of introductions between 299 and 818.
Our results suggest a large number of unobserved MERS primary cases.
Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
In these analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).

@@

% \begin{figure}[h]
% \centering
%
% 	\caption{\textbf{Pairwise co-occurence matrix for human clusters.}
% Heatmap shows the posterior probability that a pair of tips from trees of genomic fragments fall within the same clade - tips from fragment 1 are on the x axis, tips from fragment 2 are on the y axis.
% Tips are ordered by their appearance in tree of genome fragment 2 (positions from nucleotide 21000 onwards) reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
% Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
% 	}
% 	\label{coocurrence}
% \end{figure}
\begin{figure}%
    \centering
    \includegraphics[width=0.95\textwidth]{figures/mers_recombinant_features.png}
    \caption{\textbf{Recombinant features of MERS-CoV phylogenies.}
    A) Heatmap showing the posterior probability that viruses from trees of different genomic fragments belong to the same introduction event.
    Genomic fragment 1 represents nucleotide positions 1 to 21000 and genomic fragment 2 represents nucleotide positions 21001 to 29364.
    The same set of viruses are arrayed on the $x$-axis and on the $y$-axis; the $x$-axis shows identity of these viruses along genomic fragment 1 and the $y$-axis shows identity of these viruses along genomic fragment 2.
    Viruses are ordered by their appearance in tree of genomic fragment 2 reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
    Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
    B) Phylogenies derived from genomic fragment 1 (left) and genomic fragment 2 (right), reduced to just the human tips.
  	Identical taxa are connected via coloured lines with colour indicating vertical position in the tree on the right.
  	Branches are coloured by inferred ancestral host state (human in blue, camel in orange).
  	Human clusters in blue change phylogenetic positions between the trees together, with little incongruence within clusters.
  	This is evidence for recombinant viruses generated in the camel reservoir entering human populations.
    }
    \label{recombination}
\end{figure}

@@

@@

\subsection*{MERS-CoV epidemiology}
%\lmc{Detecting?}
In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula by using sequence data collected from both hosts (174 from humans and 100 from camels).
We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV genomic data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (Figures \ref{mcc}, \ref{exploded}).
This inference succeeds under different choices of priors for unknown demographic parameters (Figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (Figure \ref{mers_epi}).
From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (Figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (Figure \ref{mers_epi}).
We also looked at potential seasonality in MERS-CoV spillover into humans.
Our analyses indicated a period of three months where the odds of a sequenced spillover event are increased, with timing consistent with an enzootic amongst camel calves (Figure \ref{seasonality}).
As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (Figure \ref{mers_epi}).

Strong population structure in viruses often arises through limited gene flow, either due to geography \citep{dudas_virus_2017}, ecology \citep{smith_dating_2009} or evolutionary forces \citep{turner_genomic_2005,dudas_reassortment_2015}.
On a smaller scale population structure can unveil important details about transmission patterns, such as identifying reservoirs and understanding spill over trends and risk, much as we have done here.
When properly understood naturally arising barriers to gene flow can be exploited for more efficient disease control and prevention, as well as risk management.
%\tbc{This seems way overly broad to me. I much preferred the Introductions take on genomic epi revealing important population structure. Just revise to keep this focused on pathogen transmission. Identification of reservoirs, source-sink, all very important.}

\subsection*{Transmissibility differences between zoonoses and pandemics}
Severe acute respiratory syndrome (SARS) coronavirus, a Betacoronavirus like MERS-CoV, caused a serious epidemic in humans in 2003, with over 8000 cases and nearly 800 deaths.
Since MERS-CoV was also able to cause significant pathogenicity in the human host it was inevitable that parallels would be drawn between MERS-CoV and SARS-CoV at the time of MERS discovery in 2012.
Although we describe the epidemiology of MERS-CoV from sequence data, indications that MERS-CoV has poor capacity to spread human-to-human existed prior to any sequence data.
SARS-CoV swept through the world in a short period of time, but MERS cases trickled slowly and were restricted to the Arabian Peninsula or resulted in self-limiting outbreaks outside of the region, a pattern strongly indicative of repeat zoonotic spillover.
Infectious disease surveillance and control measures remain limited, so much like the SARS epidemic in 2003 or the H1N1 pandemic in 2009, zoonotic pathogens with $R_{0}>1.0$ are probably going to be discovered after spreading beyond the original location of spillover.
Even though our results show that MERS-CoV does not appear to present a serious global threat, we would like to highlight that its epidemiology is nonetheless concerning.
%\tbc{I like the basic approach, this last segue could be cleaner however.}

@@

\section*{Methods}
\subsection*{Sequence data}
All MERS-CoV sequences were downloaded from GenBank and accession numbers are given in Table \ref{sequences}.
Sequences without accessions were kindly shared by Ali M. Somily, Mazin Barry, Sarah S. Al Subaie, Abdulaziz A. BinSaeed, Fahad A. Alzamil, Waleed Zaher, Theeb Al Qahtani, Khaldoon Al Jerian, Scott J.N. McNabb, Imad A. Al-Jahdali, Ahmed M. Alotaibi, Nahid A. Batarfi, Matthew Cotten, Simon J. Watson, Spela Binter, and Paul Kellam prior to publication.
Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
Only sequences covering at least 50\% of MERS-CoV genome were kept.
Sequences were annotated with available collection dates and hosts, designated as camel or human and aligned with MAFFT \citep{katoh_mafft_2013} and edited manually.
Protein coding sequences were extracted and concatenated, reducing alignment length from 30130 down to 29364.
The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections (Table \ref{sequences}).

@@

MultiTypeTree module \citep{vaughan_efficient_2014} was used in BEAUti v2.4.3 \citep{bouckaert_beast_2014} to specify a structured coalescent model with two demes -- humans and camels.
This model estimates 4 parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
All analyses were run across for 200 million states across ten independent Markov chains (MCMC runs).
%\tbc{What was burn-in? How many steps in the MCMC? How was convergence assessed? How were chains combined?}
Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
Three chains out of ten did not converge and were discarded altogether.
To confirm that demographic inference was derived from the data rather than the prior we set up an additional ten independent analyses where the mean of the exponential distribution prior for migration rate was set to 10.0.
For these analyses two chains did not converge.

As a secondary analysis, we also considered a scenario where alignments split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
All analyses were set to run for 200 million states, subsampling every 20 000 states.
Chains not converging after 200 million states were discarded.
20\% of the states were discarded as burn-in, convergence assessed with Tracer 1.6 and combined with LogCombiner.
Three chains out of ten did not converge.

\subsection*{Demographic inference of MERS-CoV in the reservoir}
In order to infer the demographic history of MERS-CoV in camels we used the results of structured coalescent analyses to identify introductions of the virus into humans.
The oldest sequence from each cluster introduced into humans was used to represent a random draw from the diversity of MERS-CoV circulating in camels.
These sequences were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity (N$_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.%\lmc{already reported above. Why not remove?} \gdc{Never hurts to repeat.}
Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.

@@

  116 & Al-Hasa\_7\_2013 & KF600623, KF600655 & human & 2013-05-01 \\
  117 & Al-Hasa\_8\_2013 & KF600618, KF600626, KF600635, KF600638 & human & 2013-05-01 \\

@@

  119 & Al-Hasa\_11\_2013 & KF600629, KF600636, KF600646 & human & 2013-05-03 \\

@@

  121 & Al-Hasa\_14\_2013 & KF600615, KF600643 & human & 2013-05-08 \\

@@

  126 & Al-Hasa\_23\_2013 & KJ156860, KJ156894, KJ156929, KJ156923, KJ156862 & human & 2013-05-13 \\

@@

  134 & Taif\_2\_2013 & KJ156896, KJ156876 & human & 2013-06-12 \\
  135 & Taif\_3\_2013 & KJ156938, KJ156897, KJ156922, KJ156868, KJ156921, KJ156915,
KJ156906 & human & 2013-06-13 \\
  136 & Al-Hasa\_26\_2013 & KJ156882, KJ156941, KJ156872 & human & 2013-06-18 \\
  137 & Al-Hasa\_27\_2013 & KJ156943, KJ156939 & human & 2013-06-19 \\
  138 & Al-Hasa\_28\_2013 & KJ156887, KJ156940, KJ156889, KJ156893, KJ156884, KJ156930,
KJ156928, KJ156909 & human & 2013-06-22 \\
  139 & Riyadh\_6\_2013 & KJ156879, KJ156947, KJ156890, KJ156908, KJ156927 & human & 2013-07-02 \\

@@

  141 & Riyadh\_7\_2013 & KJ156937, KJ156905 & human & 2013-07-15 \\
  142 & Riyadh\_8\_2013 & KJ156880, KJ156942 & human & 2013-07-17 \\

@@

  145 & Asir\_2\_2013 & KJ156863, KJ156899, KJ156912, KJ156900, KJ156898, KJ156945,
KJ156932 & human & 2013-08-05 \\
  146 & Riyadh\_11\_2013 & KJ156946, KJ156911 & human & 2013-08-06 \\
  147 & Riyadh\_12\_2013 & KJ156926, KJ156901 & human & 2013-08-08 \\
  148 & Riyadh\_13\_2013 & KJ156888, KJ156873 & human & 2013-08-13 \\

@@

  150 & Hafr-Al-Batin\_4\_2013 & KJ156931, KJ156895, KJ156864, KJ156861 & human & 2013-08-25 \\
  151 & Hafr-Al-Batin\_5\_2013 & KJ156951, KJ156924, KJ156954, KJ156913 & human & 2013-08-25 \\
  152 & Riyadh\_17\_2013 & KJ156918, KJ156920, KJ156865 & human & 2013-08-26 \\

@@

  154 & Riyadh\_10\_2013 & KJ156891, KJ156936, KJ156907 & human & 2013-09-05 \\
  155 & Madinah\_3b\_2013 & KJ156950, KJ156916 & human & 2013-09-11 \\

@@

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_prior.png}
	\caption{\textbf{Result are robust to prior choice.}
Negligible flow of MERS-CoV lineages from humans into camels is recovered regardless of prior choice.
Plots show the 95\% highest posterior density for the estimated migration rate from the human deme into the camel deme (blue) and \textit{vice versa} (orange).
Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (bottom) or 10.0 (top).
	}
	\label{prior}
\end{figure}
